---
date: "September 2022"
output:
  html_document
---

```{r setup, include=FALSE}

pacman::p_unload(pacman::p_loaded(), character.only = TRUE)
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(fig.align = "center")

default.par <- par(no.readonly = T)

# function to generate column separated numbers
n.format <- function(x) {
  formatC(x, format = "d", big.mark = ",")
}

# set to English
Sys.setlocale("LC_TIME", "C")
```



The [**autoregressive integrated moving average model**](https://en.wikipedia.org/wiki/Autoregressive_integrated_moving_average){target="_blank"}, or ARIMA model, is a broadening of the class of ARMA models in order to include differencing. 
The word *integrated* is due to the fact that the stationary model is fitted based on the differenced data. 


$$y'_t = \phi_1y'_{t-1}+...+\phi_p y'_{t-p}+\theta_iw_{t-1}+...+\theta_qw_{t-q},$$
where $y"_t$ is the differenced series.

An ARIMA process of order $p$ for the AR part, order $q$ for the MA part and differences $d$ is denoted as ARIMA($p, d, q$), where

$p-$ order of the autoregressive part,   
$d-$ degree of first differencing involved and  
$q-$ order of the moving average part.   


The basic idea is that if differencing the data at some order $d$ produces an ARMA process, then the original process is said to be ARIMA. Often, a single difference will suffice to yield a stationary series. 


Many of the models we have already discussed are special cases of the ARIMA model.

$$
\begin{array}{ll}
\hline
\text{White noise} & \text{ARIMA(0,0,0)} \\
\text{Random walk} & \text{ARIMA(0,1,0) with no constant}\\
\text{Random walk with drift} & \text{ARIMA(0,1,0) with a constant} \\
\text{Autoregression} & \text{ARIMA(}p\text{,0,0)} \\
\text{Moving average} & \text{ARIMA(0,0,}q\text{)}
\end{array}
$$

```{r, echo = F}
library("htmltools")
includeHTML("https://userpage.fu-berlin.de/soga/soga-r/citation/citation.html")
```

