---
date: "September 2022"
output:
  html_document
---

```{r setup, include=FALSE}
pacman::p_unload(pacman::p_loaded(), character.only = TRUE)
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(fig.align = "center")

default.par <- par(no.readonly = T)

# function to generate column separated numbers
n.format <- function(x) {
  formatC(x, format = "d", big.mark = ",")
}

# set to English
Sys.setlocale("LC_TIME", "C")

```




The autoregressive process of order $p$ is denoted AR($p$), and defined by

$$y_t = \phi_1y_{t-1}+\phi_2y_{t-2}...+\phi_py_{t-p}+w_t\text{,}$$

which can be rewritten as

$$y_t = \sum_{i=1}^p\phi_iy_{t-i}+ w_t\text{,}$$

where $\phi_1, \phi_2,..., \phi_p$ are fixed constants and $w_t$ is a random variable with mean 0 and variance $\sigma^2$. 

If the mean, $\mu$, of $y_t$ is not zero, we replace $y_t$ by $y_t-\mu$:

$$y_t-\mu = \sum_{i=1}^p\phi_i(y_{t-i}-\mu)+ w_t$$

Autoregressive processes have a natural interpretation as the next value observed is a slight perturbation of the most recent observations or, in other words, the current value of the series is linearly dependent upon its previous values, with some random error. This model is called an *autoregressive* (AR) model since $y_t$ is regressed on
itself. 

***

The special case of $p = 1$, the first-order process, is also known as [Markov process](https://en.wikipedia.org/wiki/Markov_chain){target="_blank"}.

We generate two AR(1) series of the form

$y_t = \phi_1 y_{t-1}+w_t\text{,}$

where once $\phi_1 = 0.6$ and the other time $\phi_1 = -0.6$.

```{r}
set.seed(250)

ar1_1 <- arima.sim(list(order = c(1, 0, 0), ar = 0.6), n = 100)
ar1_2 <- arima.sim(list(order = c(1, 0, 0), ar = -0.6), n = 100)
```

We plot the two series using the `autolplot()` function.


```{r, message=FALSE, warning=FALSE}
library(ggfortify)
library(gridExtra)

p1 <- autoplot(ar1_1,
  ylab = "y",
  main = (expression(AR(1) ~ ~ ~ phi == 0.6))
)

p2 <- autoplot(ar1_2,
  ylab = "y",
  main = (expression(AR(1) ~ ~ ~ phi == -0.6))
)

grid.arrange(p1, p2, ncol = 1)
```

For an AR(1) model the current value $y_t$ of the time series is a function of the last preceding value of the series. We may review the correlational structure by plotting the correlogram. 


```{r, message=FALSE, warning=FALSE}
p1 <- autoplot(acf(ar1_1, plot = FALSE)) +
  ggtitle(expression("Serial Correlation " * AR(1) ~ ~ ~ phi == 0.6)) +
  xlim(1, 20) +
  ylim(-0.5, 0.5)

p2 <- autoplot(acf(ar1_2, plot = FALSE)) +
  ggtitle(expression("Serial Correlation " * AR(1) ~ ~ ~ phi == -0.6)) +
  xlim(1, 20) +
  ylim(-0.5, 0.5)

grid.arrange(p1, p2, ncol = 1)
```


The autocovariance function for the AR series decays exponentially (for $\phi >0$), with a possible sinusoidal variation superimposed (for $\phi <0$). Thus, the correlation drops off faster the greater the value of $k$.

***

For an AR(2) model, which can be written as 

$y_t = \phi_1 y_{t-1}+\phi_2 y_{t-2}+w_t\text{,}$

the current value $y_t$ of the time series is a function of the last two preceding values of the series.

Let us generate four different AR(2) series to review their autocorrelational structures. 

* AR(2) model I: $\phi_1 = 0.5$  and  $\phi_2 = 0.3$  
* AR(2) model II: $\phi_1 = -0.5$  and  $\phi_2 = 0.3$  
* AR(2) model III: $\phi_1 = 1$  and  $\phi_2 = -0.5$  
* AR(2) model IV: $\phi_1 = -0.5$  and  $\phi_2 = -0.3$.  


We make use of the `arima.sim()` function  to generate these four AR(2) models and plot the correlogram by using the `autoplot()` function in combination with the `acf()` function.


```{r}
ar2_I <- arima.sim(list(
  order = c(2, 0, 0),
  ar = c(0.5, 0.3)
), n = 100)

ar2_II <- arima.sim(list(
  order = c(2, 0, 0),
  ar = c(-0.5, 0.3)
), n = 100)

ar2_III <- arima.sim(list(
  order = c(2, 0, 0),
  ar = c(1, -0.5)
), n = 100)

ar2_IV <- arima.sim(list(
  order = c(2, 0, 0),
  ar = c(-0.5, -0.3)
), n = 100)
```


```{r}
p1 <- autoplot(acf(ar2_I, plot = FALSE, lag.max = 15)) +
  ggtitle(expression(AR(2) ~ ~ ~ phi[1] == 0.5 * ~ ~ phi[2] == 0.3))

p2 <- autoplot(acf(ar2_II, plot = FALSE, lag.max = 15)) +
  ggtitle(expression(AR(2) ~ ~ ~ phi[1] == -0.5 * ~ ~ phi[2] == 0.3))

p3 <- autoplot(acf(ar2_III, plot = FALSE, lag.max = 15)) +
  ggtitle(expression(AR(2) ~ ~ ~ phi[1] == 1 * ~ ~ phi[2] == -0.5))

p4 <- autoplot(acf(ar2_IV, plot = FALSE, lag.max = 15)) +
  ggtitle(expression(AR(2) ~ ~ ~ phi[1] == -0.5 * ~ ~ phi[2] == -0.3))

grid.arrange(p1, p3, p2, p4, ncol = 2)
```

The autocovariance functions exhibit exponential decay, with a possible sinusoidal variation superimposed.



```{r, echo = F}
library("htmltools")
includeHTML("https://userpage.fu-berlin.de/soga/soga-r/citation/citation.html")
```

