---
date: "September 2022"
output:
  html_document
---

```{r setup, include=FALSE}

pacman::p_unload(pacman::p_loaded(), character.only = TRUE)
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(fig.align = "center")

default.par <- par(no.readonly = T)

# function to generate column separated numbers
n.format <- function(x) {
  formatC(x, format = "d", big.mark = ",")
}

# set to English
Sys.setlocale("LC_TIME", "C")

```



A model for analyzing a series, which is the sum of a deterministic *trend* series and a stationary *noise* series, is the so called **[random walk](https://en.wikipedia.org/wiki/Random_walk){target="_blank"} with drift model**, given by

$$y_t = \delta+y_{t-1}+w_t,$$

for $t = 1, 2,...$, with initial condition $y_0 = 0$, where $w_t$ is white noise. The constant $\delta$ is called the drift, and if $\delta = 0$, then it is simply called *random walk*.

The equation from above may be rewritten as

$$y_t = \delta t + \sum_{j=1}^t w_j\text{,}$$

for $t = 1,2,...$.

If $\delta = 0$, the value of the time series at time $t$ is equal to the value of the series at time $t-1$ plus a completely random movement determined by $w_t$. 

$$\Delta y_t = y_t-y_{t-1} = w_t$$

The $w_{t}$ values can be interpreted as independent “shocks” which perturb the current state $y_t$ in the amount of $w_{t}$ to produce a new state $y_{t+1}$.

The graph below shows a random walk and a random walk with drift ($\delta = 0.25$) generated with R.

```{r}
set.seed(250)

## random walk
w <- rnorm(n = 200, mean = 0, sd = 1)
y <- cumsum(w)

## random walk with drift
wd <- w + .25
yd <- cumsum(wd)

## plotting
plot.ts(yd,
  ylim = c(-5, 55),
  main = expression(paste("Random walk with " * sigma[w] * " = 1")),
  col = "blue"
)

lines(y, col = "red")

legend("topleft",
  legend = c(
    expression(paste(delta * " = 0.25")),
    expression(paste(delta * " = 0"))
  ),
  col = c("blue", "red"),
  lty = 1
)
```

The autocorrelation function of this model type shows typically a very slow decay. 

```{r, message=FALSE, warning=FALSE}
library(ggfortify)
library(gridExtra)

p1 <- autoplot(acf(y, plot = FALSE)) +
  ggtitle("Serial Correlation of a random walk")

p2 <- autoplot(acf(yd, plot = FALSE)) +
  ggtitle("Serial Correlation of a random walk with drift")
grid.arrange(p1, p2, ncol = 1)
```


```{r, echo = F}
library("htmltools")
includeHTML("https://userpage.fu-berlin.de/soga/soga-r/citation/citation.html")
```

